Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-07-27 Manuscript/00-main.Rmd
Finish revising the conclusion part

...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`. It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323).

Analogously, use Rmarkdown to produce tables as usual:

>>>>

<!-- ...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`. It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323). -->

<!-- Analogously, use Rmarkdown to produce tables as usual: -->

### 2021-07-27 Manuscript/SS_GAM.tex
Finish revising the conclusion part

@@

\includegraphics{SS_GAM_files/figure-latex/unnamed-chunk-10-1.pdf}

@@

Remaining issue: 1. Degree of freedom has yet to closely studied. As an
direct impact, the functional selection can be improved when effective
degree of freedom are more carefully studied 2. While the proposed model
addresses a great deal of analytic problem, analyzing the time-to-event
outcome remains unsolved. Our current effort resides on expanding. An
naive approach could be convert an time-to-event outcome to a poisson
outcome following xxx. Hence, the . However, it would be more efficient
to , via penalized pseudo likelihood function, which is the current
effort resides. 2. Future exapansion, as we were analyzing the
metabolites data, we observed that metabolites pathway infromation can
be useful in the predictive and integrative model can be developed based
on the proposed framework. As an naive solution, the selection indicator
of linear effect of the spline terms of all metabolites on the same
pathway can be set to share the same hyper prior such that the group
selection can be modestly achieved. Meanwhile further advancedment of
the methodology regarding to the integrative analysis can consider
setting the hypor priors of the selection indivator of the non-linear
components conditioning on the probability of inclusion of the linear
effect. Such structure is more sensible analytically.

Meanwhile, we specify that \(\gamma_{j}^0\) and \(\gamma_{j}^{pen}\) are
independently distributed for analytic simplicity. However, this set-up
could possibility result in a situation that is not theoretically
possible: the non-linear component is selected, but the linear component
is not. This can be addressed analytically by forcing to include the
linear component when non-linear component is included. Another possible
solution is to impose an dependent structure of \(\gamma_{j}^{pen}\) on
\(\gamma_{j^{0}}\), i.e.~\(\gamma_j^{pen}|\gamma_{j}^{0}, \theta_j\).

\hypertarget{discussion-1}{%
\subsection{Discussion}\label{discussion-1}}

In this chapter, we propose the spike-and-slab generalized additive
models for high-dimensional analysis.Two fast computing algorithms,
EM-IRLS and EM-coordinate descent are developed for expedited model
fitting. EM-IRLS provides uncertainty measures, while EM-coordinate
descent algorithm are extremely fast for model fitting. In the
simulation studies, we shows the proposed model outperforms the state of
the art GAM model, \texttt{mgcv} in prediction and function inference.
Moreover, due to the model selection nature of the spike-and-slab prior,
the proposed model can identify predictors that have large effects on
the outcome. In addition, the model can be used to answer the question
if non-linear effect are necessary.

The ``all-in-all-out'' set-up was criticized for its inability to
answering an important research question, if the effects are non-linear
or not. On the other hand, the model can be much more flexible by
allowing each dimension of the penalized space to have an independent
indicator, i.e.~\(\gamma^{pen}_{jk}\) of \(\beta^{pen}_{jk}\) for
\(k = 1\dots, K_j\). This prior set-up generalized to the group
spike-and-slab. The additional flexibility doesn't not necessarily grand
the model neither theoretical justification nor analytic simplicity. The
spline penalty assumes a common smooth shrinkage \(\lambda_j\) applied
to all coefficients of the penalized space, which contradicts to the
model setup. Besides, having multiple indicator variables complicates
the decision of if non-linear effects are necessary.

In the manuscript, we describe the implementation of ssGAM-NM using
EM-IRLS. Nevertheless, EM-IRLS are not limited to fitting ssGAM-NM.
Other spike-and-slab prior can be fitted with EM-IRLS as well,
e.g.~mixture t-distribution, and mixture double exponential
distributions. \citet{Yi2012} demonstrated that t prior and double
exponential prior can be seen hierarchical normal priors, \[
\begin{aligned}
\beta_i \sim N(0, \tau^2_i),\\
\tau^2_i \sim Inv-\chi^2(v, s^2_j) \text{ or } \tau^2_i \sim Gamma(1, s^2/2)
\end{aligned}
\] with \(\tau^2\) integrated out analytically. Hence, the EM-IRLS can
be easily applied to those two additional mixture priors. One of the
advantage of using mixture \(t\) prior over mixture normal prior is that
\(t\) distributions with low degree of freedom, e.g.~\(v=1\), have
heavier tail than the normal distribution, which is also a special case
of \(t\) distribution when \(v=\infty\). The heavier tails help to
obtain posteriors that concentrated around the true nonzero coefficients
at optimal rates. \citep{Rockova2014a}

\ldots and can be referenced (Figure \ref{fig:plot}) by including the
\texttt{\textbackslash{}\textbackslash{}label\{\}} tag in the
\texttt{fig.cap} attribute of the R chunk:
\texttt{fig.cap\ =\ "Fancy\ Caption\textbackslash{}\textbackslash{}label\{fig:plot\}"}.
It is a quirky hack at the moment, see
\href{https://github.com/yihui/knitr/issues/323}{here}.

Analogously, use Rmarkdown to produce tables as usual:

\hypertarget{cross-referencing}{%
\subsection{Cross-referencing}\label{cross-referencing}}

The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
for figures, tables, equations, etc., is encouraged (using
\texttt{{[}@\textless{}name\textgreater{}{]}}, equivalent of
\texttt{\textbackslash{}ref\{\textless{}name\textgreater{}\}} and
\texttt{\textbackslash{}label\{\textless{}name\textgreater{}\}}). That
works well for citations in Rmarkdown, not so well for figures and
tables. In that case, it is possible to revert to standard
\LaTeX syntax.

>>>>

\includegraphics{SS_GAM_files/figure-latex/unnamed-chunk-7-1.pdf}

@@

@@

In the paper, we described a novel generalized additive model using
Bayesian hierarchical priors, particularly the proposed spike-and-slab
spline prior for bi-level functional selection. Meanwhile, we introduced
two optimization based algorithms for model fitting. The algorithms can
be easily scale up to address high-dimensional data analysis in a
computational efficient manner, but also provide naive inferential
supports for many commonly used hierarchical priors,
e.g.~t-distribution, double exponential distribution, mixture t
distribution, and mixture double exponential distribution, in the
analyses that the sparse assumption is weak or not necessary. Via
simulations, we demonstrated that the proposed model provides as good,
if not better, prediction performance compared to some state-of-the-art
non-linear modeling devices.

The proposed model shares many commonality with an high-dimensional
Bayesian GAM, SB-GAM \citep{Bai2021}, independently developed around the
same time of this work. Both frameworks emphasize computational
efficiency by deploying group spike-and-slab lasso type of priors and
optimization-based fast and scalable algorithms. Bai provided the
theoretical proof for the consistency of variable selection using group
spike-and-slab lasso prior. Nevertheless, SB-GAM fails to address the
bi-level functional selection and the model inference. The proposed
model provides solutions to these remaining questions while maintaining
the same level of prediction accuracy. Moreover, the proposed model
renders additional flexibility to model specification, allowing various
choices of smoothing functions and degrees of freedom.

For transnational science purpose, we implemented the proposed model in
a R package \texttt{BHAM}, deposited at
\url{https://github.com/boyiguo1/BHAM}. To maximize the flexibility of
smoothing function specification, we deploy the same programming grammar
as in the state-of-the-art package \texttt{mgcv}, in contrast to
previous tools where smoothing functions are limited to the default
ones. Ancillary functions are provided for model specification in
high-dimensional settings, curve plotting and functional selection.

There are some improvements possible for the proposed models. First of
all, the proposed model achieves a bi-level selection via the two-part
spike-and-slab spline prior. Nevertheless, this set-up could result in a
situation that is not theoretically sound: the non-linear component is
selected, but the linear component is not. We currently address it
analytically by forcing to include the linear component when non-linear
component is included. Another possible solution is to impose an
dependent structure of \(\gamma_{j}^{pen}\) on \(\gamma_{j^{0}}\),
i.e.~\(\gamma_j^{pen}|\gamma_{j}^{0}, \theta_j\). Secondly, while the
proposed model addresses a great deal of analytic problem, analyzing the
time-to-event outcome remains unsolved. Our current effort resides on
expanding GAM to Cox proportional hazard models. An naive approach would
be convert an time-to-event outcome to a poisson outcome following
Whitehead \citep{Whitehead1980}. However, it would be more efficient to
directly fitting Cox models via penalized pseudo likelihood function.
Last but not least, we plan to expand the proposed model to integrate
external biology information. The main motivation for the integrative
model is that biologically informed groupping of weak effects increases
the power of detecting true associations between features and the
outcome \citep{Peterson2016}, and stabilizes the analysis results for
reproducibility purpose. Such integration can be achieved by setting up
a structural hyperprior on the inclusion indicator of the smoothing
function null space \(\boldsymbol{\gamma}^0\). The similar strategy has
been used in Ferrari and Dunson \citep{Ferrari2020}.
